---
title: "Differential_expression"
author: "Prymidis Dimitrios"
date: "11/30/2020"
output: 
  html_document:
    toc: true
    number_sections: true
---

In this R.Markdown we are going to do differential gene analysis using welch's t-test.
2 Groups: Group 1: "trunc", Group 2: "no_trunc"
Per gene: Welch's t-test on expression level ad RNA Ranks
The genes we are interested are the following (11):
AXIN2, GUCA2A, HNRNPA0, NOLC1, RNF43, SNRPA1, SOX9, TCF7, VILL, ZNHIT2, ZNRF3


Here we are using FPKM values

#  Load libraries 
```{r setup, include=FALSE, echo=TRUE, results = 'hide'}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE, message = FALSE
)

library(tidyverse)
library(plotly)
library(IRdisplay)
library(ggplot2)
library(ggpubr)
library(rstatix)

```

Set working directory
```{r}
setwd("~/Documents/Master/internship/APC/APC_Project")
```

# load sample data
Group 1: "trunc", Group 2: "no_trunc"
```{r load sample data}

samp_to_class <- read.csv("./1_trunc_samples.filtered") # samples with 0 or 1 truncating mutation

samp_to_class$Class[samp_to_class$Class != 0] <- "trunc" 
samp_to_class$Class[samp_to_class$Class == 0] <- "no_trunc"
tsb <- samp_to_class[,1] # make tumor sample barcode rownames
rownames(samp_to_class) <- tsb
table(samp_to_class$Class)
samp_to_class$Class <- as.factor(samp_to_class$Class) # Make class column a factor
samp_to_class$Class <- relevel(samp_to_class$Class, ref =  "no_trunc") # make no_trunc the reference level
```

# Load count data

```{r}

# Load all gene expression countdata
CC <- as.data.frame(read.csv("./Data/Colorectal/Gene_expression/CRC_Gene_expression_all_Counts"))
names(CC)[1] <- "Tumor_Sample_Barcode"  # rename first column to be Tumor Sample Barcode
Counts<- CC

# Take the correct samples from Gene expression data
Counts$Tumor_Sample_Barcode <- sapply(Counts$Tumor_Sample_Barcode, function(x){substring(x, 1, 15)})
Counts <- Counts[Counts$Tumor_Sample_Barcode %in% samp_to_class$Tumor_Sample_Barcode,]
set.seed(123)
Counts <-Counts %>% distinct(Tumor_Sample_Barcode, .keep_all = TRUE) ### take unique rows 
rownames(Counts) <- Counts$Tumor_Sample_Barcode # make tumor sample barcode rownames
Counts$Tumor_Sample_Barcode <- NULL # and delete the column

# Remove columns with low counts and or zero variance variables
temp <- Counts[, - as.numeric(which(apply(Counts, 2, var) == 0))] # Remove columns with zero variance 
temp2<- temp[,colSums(temp)>10] # Remove column with low counts 
Counts <- NULL
Counts<- temp2

# Merge classes with Gene expression data
Counts$Tumor_Sample_Barcode <- rownames(Counts)
Counts <- merge(samp_to_class, Counts, by = 'Tumor_Sample_Barcode')
rownames(Counts) <- Counts$Tumor_Sample_Barcode
Counts$Tumor_Sample_Barcode <- NULL
```
There are no columns with zero variance nor columns with only zeros

# z-normalization. (Center and scale) and log scale

Using z-normalized data didnt change any of the following results
For this Rmd we are using the raw counts

```{r}
Counts.scaled <- Counts
Counts.scaled[, -c(1)] <- scale(Counts.scaled[, -c(1)], center = TRUE, scale = TRUE)
Counts.log <- Counts
Counts.log[, -c(1)] <- log(Counts.log[, -c(1)]+1) # take the log base e

#Counts<-Counts.log 
```

# Violin Plots


```{r}
## 1st way
mean_sd <- function(x) {
   m <- mean(x)
   ymin <- m-sd(x)
   ymax <- m+sd(x)
   return(c(y=m,ymin=ymin,ymax=ymax))}

#pAXIN2 <- ggplot(Counts.log, aes(x=Class, y=AXIN2)) + geom_violin() + stat_summary(fun.data=mean_sd)


########## 2nd way
# errbar_lims <- group_by(Counts.log, Class) %>% summarize(mean=mean(AXIN2), se=sd(AXIN2)/sqrt(n()),   upper=mean+(2*se), lower=mean-(2*se))
# pAXIN2 <- ggplot() + geom_violin(data=Counts.log, aes(x=Class, y=AXIN2)) + geom_point(data=errbar_lims, aes(x=Class, y=mean), size=3) +
 # geom_errorbar(aes(x=errbar_lims$Class, ymax=errbar_lims$upper,   ymin=errbar_lims$lower), stat='identity', width=.25) + theme_minimal()
###########
# We are going to use the first way

pAXIN2 <- ggplot(Counts, aes(x=Class, y=AXIN2)) + 
  geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pGUCA2A <- ggplot(Counts, aes(x=Class, y=GUCA2A)) + 
geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pHNRNPA0 <- ggplot(Counts, aes(x=Class, y=HNRNPA0)) + 
geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pNOLC1 <- ggplot(Counts, aes(x=Class, y=NOLC1)) + 
 geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pRNF43 <- ggplot(Counts, aes(x=Class, y=RNF43)) + 
 geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pSNRPA1 <- ggplot(Counts, aes(x=Class, y=SNRPA1 )) + 
  geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pSOX9 <- ggplot(Counts, aes(x=Class, y=SOX9)) + 
  geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pTCF7 <- ggplot(Counts, aes(x=Class, y=TCF7 )) + 
 geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pVILL <- ggplot(Counts, aes(x=Class, y=VILL )) + 
  geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pZNHIT2 <- ggplot(Counts, aes(x=Class, y=ZNHIT2 )) + 
 geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pZNRF3 <- ggplot(Counts, aes(x=Class, y=ZNRF3 )) + 
geom_violin() + stat_summary(fun.data=mean_sd, color = "red")

ggarrange(pAXIN2, pGUCA2A, pHNRNPA0, pNOLC1, pRNF43, pSNRPA1, pSOX9, pTCF7, pVILL, pZNHIT2, pZNRF3, ncol = 4, nrow = 3)
```


# AXIN2 Plots for data visualisation

````{r}
summary(Counts %>% filter(Class == "no_trunc") %>% .$AXIN2)
summary(Counts %>% filter(Class == "trunc") %>% .$AXIN2)

p1 <-Counts %>% filter(Class == "no_trunc") %>% ggplot(aes(AXIN2)) + 
        geom_histogram(fill = "white", color = "grey30")

p2 <-Counts %>% filter(Class == "trunc") %>% ggplot(aes(AXIN2)) + 
        geom_histogram(fill = "white", color = "grey30")

ggarrange(p1, p2, ncol = 2)

ggplot(Counts, aes(Class, AXIN2)) +
        geom_boxplot()
```


The left plots are the "no_trunc" and the right are the "trunc"
 
```{r}
Counts$Class.x <- NULL
c01<-  Counts %>% filter(Class.y != "2")
c02<-  Counts %>% filter(Class.y != "1") 
```
# t-test on the Counts

The genes we are interested are the following (11):
AXIN2, GUCA2A, HNRNPA0, NOLC1, RNF43, SNRPA1, SOX9, TCF7, VILL, ZNHIT2, ZNRF3

```{r}
t.test(AXIN2 ~ Class.y, data = c02)
t.test(GUCA2A ~ Class.y, data = c02)
t.test(HNRNPA0 ~ Class.y, data = c02)
t.test(NOLC1 ~ Class.y, data = c02)
t.test(RNF43 ~ Class.y, data = c02)
t.test(SNRPA1 ~ Class.y, data = c02)
t.test(SOX9 ~ Class.y, data = c02)
t.test(TCF7 ~ Class.y, data = c02)
t.test(VILL ~ Class.y, data = c02)
t.test(ZNHIT2 ~ Class.y, data = c02)
t.test(ZNRF3 ~ Class.y, data = c02)

```

# RANK
Calculate the rank of the samples for the genes we are interested in.


````{r}
rAXIN2<- Counts %>% select(AXIN2, Class) %>% arrange(desc(AXIN2)) %>% mutate(rank =row_number())
rGUCA2A<- Counts %>% select(GUCA2A, Class) %>% arrange(desc(GUCA2A)) %>% mutate(rank =row_number())
rHNRNPA0<- Counts %>% select(HNRNPA0, Class) %>% arrange(desc(HNRNPA0)) %>% mutate(rank =row_number())
rNOLC1<- Counts %>% select(NOLC1, Class) %>% arrange(desc(NOLC1)) %>% mutate(rank =row_number())
rRNF43<- Counts %>% select(RNF43, Class) %>% arrange(desc(RNF43)) %>% mutate(rank =row_number())
rSNRPA1<- Counts %>% select(SNRPA1, Class) %>% arrange(desc(SNRPA1)) %>% mutate(rank =row_number())
rSOX9<- Counts %>% select(SOX9, Class) %>% arrange(desc(SOX9)) %>% mutate(rank =row_number())
rTCF7<- Counts %>% select(TCF7, Class) %>% arrange(desc(TCF7)) %>% mutate(rank =row_number())
rVILL<- Counts %>% select(VILL, Class) %>% arrange(desc(VILL)) %>% mutate(rank =row_number())
rZNHIT2<- Counts %>% select(ZNHIT2, Class) %>% arrange(desc(ZNHIT2)) %>% mutate(rank =row_number())
rZNRF3<- Counts %>% select(ZNRF3, Class) %>% arrange(desc(ZNRF3)) %>% mutate(rank =row_number())

```


# t-test on Ranks
The genes we are interested are the following (11):
AXIN2, GUCA2A, HNRNPA0, NOLC1, RNF43, SNRPA1, SOX9, TCF7, VILL, ZNHIT2, ZNRF3 


````{r}
t.test(rank ~ Class, data = rAXIN2)
t.test(rank ~ Class, data = rGUCA2A)
t.test(rank ~ Class, data = rHNRNPA0)
t.test(rank ~ Class, data = rNOLC1)
t.test(rank ~ Class, data = rRNF43)
t.test(rank ~ Class, data = rSNRPA1)
t.test(rank ~ Class, data = rSOX9)
t.test(rank ~ Class, data = rTCF7)
t.test(rank ~ Class, data = rVILL)
t.test(rank ~ Class, data = rZNHIT2)
t.test(rank ~ Class, data = rZNRF3)
```

# Get the p-values for the Counts and for the Ranks
The genes we are interested are the following (11):
AXIN2, GUCA2A, HNRNPA0, NOLC1, RNF43, SNRPA1, SOX9, TCF7, VILL, ZNHIT2, ZNRF3 

````{r}
# Counts
cAXIN2p <- t.test(AXIN2 ~ Class, data = Counts)$p.value
cGUCA2Ap <-t.test(GUCA2A ~ Class, data = Counts)$p.value
cHNRNPA0p <-t.test(HNRNPA0 ~ Class, data = Counts)$p.value
cNOLC1p <-t.test(NOLC1 ~ Class, data = Counts)$p.value
cRNF43p <-t.test(SNRPA1 ~ Class, data = Counts)$p.value
cSNRPA1p <-t.test(SNRPA1 ~ Class, data = Counts)$p.value
cSOX9p <-t.test(TCF7 ~ Class, data = Counts)$p.value
cTCF7p <-t.test(SOX9 ~ Class, data = Counts)$p.value
cVILLp <-t.test(VILL ~ Class, data = Counts)$p.value
cZNHIT2p <-t.test(ZNHIT2 ~ Class, data = Counts)$p.value
cZNRF3p <-t.test(ZNRF3 ~ Class, data = Counts)$p.value

# Ranks
rAXIN2p <- t.test(rank ~ Class, data = rAXIN2)$p.value
rGUCA2Ap <- t.test(rank ~ Class, data = rGUCA2A)$p.value
rHNRNPA0p <-t.test(rank ~ Class, data = rHNRNPA0)$p.value
rNOLC1p <-t.test(rank ~ Class, data = rNOLC1)$p.value
rRNF43p <- t.test(rank ~ Class, data = rRNF43)$p.value
rSNRPA1p <-t.test(rank ~ Class, data = rSNRPA1)$p.value
rSOX9p <-t.test(rank ~ Class, data = rSOX9)$p.value
rTCF7p <-t.test(rank ~ Class, data = rTCF7)$p.value
rVILLp <-t.test(rank ~ Class, data = rVILL)$p.value
rZNHIT2p <-t.test(rank ~ Class, data = rZNHIT2)$p.value
rZNRF3p <- t.test(rank ~ Class, data = rZNRF3)$p.value
```

# Count and Rank p.values

Table with the p.values from Welch's t-test between  "no_trunc" and "trunc" samples using the Counts and the Ranks 


```{r}
df <- as.data.frame(matrix(0, ncol = 1, nrow = 11))
names(df)<- "Genes"             
df$Genes<- c("AXIN2", "GUCA2A", "HNRNPA0", "NOLC1", "RNF43", "SNRPA1", "SOX9", "TCF7", "VILL", "ZNHIT2", "ZNRF3")
df$Counts <- c(cAXIN2p, cGUCA2Ap, cHNRNPA0p, cNOLC1p, cRNF43p, cSNRPA1p, cSOX9p, cTCF7p, cVILLp, cZNHIT2p, cZNRF3p)
df$Ranks <- c(rAXIN2p, rGUCA2Ap, rHNRNPA0p, rNOLC1p, rRNF43p, rSNRPA1p, rSOX9p, rTCF7p, rVILLp, rZNHIT2p, rZNRF3p )

df %>% knitr::kable()
```


# Violin Plots for Counts with p.values


```{r}
ggarrange(pAXIN2 + rremove("xlab") , pGUCA2A + rremove("xlab"), pHNRNPA0 + rremove("xlab"), pNOLC1 + rremove("xlab"), pRNF43 + rremove("xlab"), pSNRPA1 + rremove("xlab"), pSOX9 + rremove("xlab"), pTCF7 + rremove("xlab"), pVILL + rremove("xlab"), pZNHIT2 + rremove("xlab"), pZNRF3 + rremove("xlab"), ncol = 4, nrow = 3, labels =  df$Counts)
```

# Violin Plots for Ranks  with p.values


```{r}
pAXIN2 <- ggplot(rAXIN2, aes(x=Class, y=rank)) + 
  geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pGUCA2A <- ggplot(rGUCA2A, aes(x=Class, y=rank)) + 
geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pHNRNPA0 <- ggplot(rHNRNPA0, aes(x=Class, y=rank)) + 
geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pNOLC1 <- ggplot(rNOLC1, aes(x=Class, y=rank)) + 
 geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pRNF43 <- ggplot(rRNF43, aes(x=Class, y=rank)) + 
 geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pSNRPA1 <- ggplot(rSNRPA1, aes(x=Class, y=rank )) + 
  geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pSOX9 <- ggplot(rSOX9, aes(x=Class, y=rank)) + 
  geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pTCF7 <- ggplot(rTCF7, aes(x=Class, y=rank )) + 
 geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pVILL <- ggplot(rVILL, aes(x=Class, y=rank )) + 
  geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pZNHIT2 <- ggplot(rZNHIT2, aes(x=Class, y=rank )) + 
 geom_violin() + stat_summary(fun.data=mean_sd, color = "red")
pZNRF3 <- ggplot(rZNRF3, aes(x=Class, y=rank )) + 
geom_violin() + stat_summary(fun.data=mean_sd, color = "red")

ggarrange(pAXIN2 + rremove("xlab") , pGUCA2A + rremove("xlab"), pHNRNPA0 + rremove("xlab"), pNOLC1 + rremove("xlab"), pRNF43 + rremove("xlab"), pSNRPA1 + rremove("xlab"), pSOX9 + rremove("xlab"), pTCF7 + rremove("xlab"), pVILL + rremove("xlab"), pZNHIT2 + rremove("xlab"), pZNRF3 + rremove("xlab"), ncol = 4, nrow = 3, labels =  df$Ranks)

```



