---
title: "PCA for samples with 1 truncating mutation"
author: "Prymidis Dimitrios"
date: "11/30/2020"
output: 
  html_document:
    toc: true
    number_sections: true
---

#  Load libraries 
```{r setup, include=FALSE, echo=TRUE, results = 'hide'}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE, message = FALSE
)

library(tidyverse)
library(factoextra)
```

Set working director
```{r}
setwd("~/Documents/Master/internship/APC/APC_Project")
```


The Class data contain the Tumor Sample Barcode of samples with no and 1 APC truncating mutation.  It also has a column "Class" which as the class of each sample (0to5, 0: samples with no APC mutation, 1: for samples with APC truncating mutation in zone 1, 2:...).

To keep this simple we take into consideration only 2 classes. 
Class 1: samples with no APC mutation (as "no_trunc") 
Class 2: samples with 1 APC truncating mutation ( as "trunc")

```{r load sample data}
samp_to_class <- read.csv("./1_trunc_samples.prefiltered") # samples with 0 or 1 truncating mutation

samp_to_class$Class[samp_to_class$Class != 0] <- "trunc" #Class 0 samples with no trunc mutation vs Class 1 samples with 1 trunc mutation
samp_to_class$Class[samp_to_class$Class == 0] <- "no_trunc"
tsb <- samp_to_class[,1] # make tumor sample barcode rownames
rownames(samp_to_class) <- tsb
samp_to_class$Class <- as.factor(samp_to_class$Class) # Make class column a factor
samp_to_class$Class <- relevel(samp_to_class$Class, ref =  "no_trunc") # make no_trunc the reference level
table(samp_to_class$Class)
```
# Remove CNA
```{r}
#CNV <- read.csv("./produced_data/APC_CNV") 
#colnames(CNV) <- c("cna", "Tumor_Sample_Barcode")
#CNV$Tumor_Sample_Barcode <- sapply(CNV$Tumor_Sample_Barcode, function(x){substring(x, 1, 15)})
#a<- CNV %>% filter(cna != 0)
#samp_to_class <- samp_to_class[!samp_to_class$Tumor_Sample_Barcode %in% a$Tumor_Sample_Barcode,]
#table(samp_to_class$Class)
```

Load Gene expression data

We remove genes with no counts and genes with zero variance

```{r PCA on all the data}

# Load all gene expression countdata
CC <- as.data.frame(read.csv("./Data/Colorectal/Gene_expression/CRC_Gene_expression_all_Counts"))
names(CC)[1] <- "Tumor_Sample_Barcode"  # rename first column to be Tumor Sample Barcode
Counts<- CC

# Take the correct samples from Gene expression data
Counts$Tumor_Sample_Barcode <- sapply(Counts$Tumor_Sample_Barcode, function(x){substring(x, 1, 15)})
Counts <- Counts[Counts$Tumor_Sample_Barcode %in% samp_to_class$Tumor_Sample_Barcode,]
set.seed(123)
Counts <-Counts %>% distinct(Tumor_Sample_Barcode, .keep_all = TRUE) ### take unique rows (it is a bit random which data it is taking)
rownames(Counts) <- Counts$Tumor_Sample_Barcode # make tumor sample barcode rownames
Counts$Tumor_Sample_Barcode <- NULL # and delete the column

# Remove columns with low counts and or zero variance variables
temp <- Counts[, - as.numeric(which(apply(Counts, 2, var) == 0))] # Remove columns with zero variance 
temp2<- temp[,colSums(temp)>10] # Remove column with low counts 
Counts <- NULL
Counts<- temp2
Counts <- t(Counts) # transpose the data frame. We want genes are rows ans samples as columns

#### Make the sample indexes the same
all(colnames(Counts) %in% rownames(samp_to_class)) # Check if there are the same samples 
all(colnames(Counts) == rownames(samp_to_class))# Check if the samples are lined up correctly
samp_to_class <- samp_to_class[match(colnames(Counts), rownames(samp_to_class)), ]
all(colnames(Counts) == rownames(samp_to_class)) # check again
```

# PCA
Another method to view the relationships between samples is principal components analysis (PCA).
Hese we use log(cpm) from our raw counts

```{r pca}
y = droplevels(samp_to_class$Class)
res.pca <-  prcomp(t(Counts), scale = TRUE)
fviz_eig(res.pca) # Plot of explained variance # In all the data we see that the first component explains less variance but the second more.
fviz_pca_ind(res.pca,
             col.ind = y, labels = FALSE)
```

Remove outliers
```{r}
samp_to_class <- samp_to_class %>% filter(samp_to_class$Tumor_Sample_Barcode != "TCGA-A6-5656-01" & samp_to_class$Tumor_Sample_Barcode != "TCGA-A6-2684-01" & samp_to_class$Tumor_Sample_Barcode != "TCGA-AZ-4323-01")# &   samp_to_class$Tumor_Sample_Barcode != "TCGA-A6-6648-01" & samp_to_class$Tumor_Sample_Barcode != "TCGA-DM-A1DB-01" &  samp_to_class$Tumor_Sample_Barcode != "TCGA-AZ-4323-01" & samp_to_class$Tumor_Sample_Barcode != "TCGA-AA-3489-01"& samp_to_class$Tumor_Sample_Barcode != "TCGA-CM-6167-01" & samp_to_class$Tumor_Sample_Barcode != "TCGA-A6-6651-01")

# save the filtered samples
samp_to_class <- read.csv("./1_trunc_samples.prefiltered") # samples with 0 or 1 truncating mutation
samp_to_class <- samp_to_class %>% filter(samp_to_class$Tumor_Sample_Barcode != "TCGA-A6-5656-01" & samp_to_class$Tumor_Sample_Barcode != "TCGA-A6-2684-01" & samp_to_class$Tumor_Sample_Barcode != "TCGA-AZ-4323-01")


write.csv(samp_to_class,"1_trunc_samples.filtered", row.names = FALSE)

```








